# Mathematical optimization {#mathematical_optimization}

- Optimization uses a rigorous *mathematical model* to determine the most efficient solution to a described problem 
- One must first identify an *objective*
- Objective is a quantitative measure of the performance
- Examples: profit, time, cost, potential energy
- In general, any quantity (or combination thereof) represented as a *single number*
```{r fig.align="center", out.width="80%", echo=FALSE}
include_figure("extreme_value_theorem.png")
```

## Applications of Optimization

- *Management* 
    - Determining product portfolios
    - Location planning
    - Investments decisions
- *Game theory* 
    - Comparing players' strategies
- *Logistics* 
    - Finding optimal routes and schedules
- *Design decisions* 
    - Constructing processes, plants and other equipment
- *Operation* 
    - Adjustment to changes in environmental conditions, production planning, control, etc.
- *Mathematical modeling* 
    - Parameter estimation
    - Model discrimination

## Optimization Problem
Optimization is the *minimization or maximization* of a function subject to (s.t.) constraints on its variables

**Notation:** General form

\begin{align*}
\min\limits_{\boldsymbol{x}} f(\boldsymbol{x}) \quad\text{s.t.}\quad & h(\boldsymbol{x}) = 0 \\
 & g(\boldsymbol{x}) \leq 0
\end{align*}

with

- $\boldsymbol{x} \in\R^n$ as the *variable*, *unknown* or *parameter*
- *Objective function* $f : D \rightarrow \R$, $D \subseteq \R^n$
- *Equality constraints* $h : D_h \rightarrow \R^l$, $D_h \subseteq \R^n$
- *Inequality constraints* $g : D_g \rightarrow \R^k$, $D_g \subseteq \R^n$

## Properties of Optimization Problems

- **Objective**: Linear, quadratic, non-linear, etc. 
- **Constraints**: Equality and inequality 
- **Variable types**: $\boldsymbol{x}$ can be continuous, integer, mixed 
- **Direction**: $\min\limits_{\vecval{x}} f(\vecval{x}) \Leftrightarrow \max\limits_{\vecval{x}} -f(\vecval{x})$ 
- **Bounds**: Lower $\vecval{x}_L \leq \vecval{x}$ or upper $\vecval{x} \leq \vecval{x}_U$ 
- **Dimension**: One dimensional if $n = 1$, or multi-dimensional if $n > 1$
- **Optima**: Isolated, local or global nature 

## Classification of Optimization Problems
- *Linear Programming (LP)* 
    - Objective function and constraints are linear
    - $\min\limits_{\vecval{x}} \vecval{c}^T \vecval{x}$ s.t. $A \vecval{x} \leq \vecval{b}$, $\vecval{x} \geq 0$
- *Quadratic Programming (QP)* 
    - Objective function is quadratic and constraints are linear
    - $\min\limits_{\vecval{x}} \vecval{x}^T Q \vecval{x}$ s.t. $A \vecval{x} \leq \vecval{b}$, $\vecval{x} \geq 0$
- *Non-Linear Programming (NLP):* objective function or at least one constraint is non-linear
- *Integer Programming (IP):* all variables are discrete
- *Mixed Integer Programming (MIP)* 
    - Continuous and discrete variables
    - Problem can be linear (MILP) or non-linear (MINLP)

## Classification of Optimization Problems
- *Dynamic Optimization:* solution is a function of time
- *Stochastic Optimization*
    - Model cannot be fully specified, but has uncertainties with confidence estimates
    - Optimize expected performance given uncertainty


### Question {-}
- What type is the following optimization problem?
$$
\max\limits_{x,y} 3x + y^2
\quad
\text{s.t.}
\quad 
x + y < 10 \text{ and } y \in \left\{ 1, 2, 4, 8 \right\}
$$
a) MIP
b) MILP
c) MINLP
- Pingo vielleicht, dunno ob das solange aktiv ist / sein soll(TODO?)

## Optimal Solution

```{r fig.align="center", out.width="80%", echo=FALSE}
include_figure("optimization_problem.png")
``` 

- $x^\ast$ is a *global minimum* if $x^\ast \in D$ and $f(x^\ast) \leq f(x) \qquad \text{for all } x \in D$

- $\rightarrow$ Global minimizers are desired, though often one has only local knowledge of $f$

### Examples of optimal solutions {-}

```{r fig.align="center", fig.cap="Isolated global solution", out.width="80%", echo=FALSE}
include_figure("isolated_global_solution.png")
``` 

```{r fig.align="center", fig.cap="One global, two local solutions", out.width="80%", echo=FALSE}
include_figure("two_local_solutions.png")
``` 

```{r fig.align="center", fig.cap="A local but no global solution", out.width="80%", echo=FALSE}
include_figure("local_non-global_solution.png")
``` 

```{r fig.align="center", fig.cap="Many non-isolated, global solutions", out.width="80%", echo=FALSE}
include_figure("non-isolated_global_solution.png")
``` 

## Optimization Procedure
*Formulation and solution* of optimization problems usually follows:

1. *Analysis of environment* to determine the variables of interest
1. Definition of optimality criteria as an *objective function* with (additional) constraints
1. Formulation as a *mathematical model* with degrees of freedom
1. *Numerical optimization* to find a solution
1. *Verification of the solution* through sensitivity analysis (with respect to the assumptions made in the problem formulation)

## Motivation

### Production Scheduling
- **Problem**
    - Given plants $i = 1, \ldots, M$ where each *manufactures* $P_i$ goods
    - Each plant has a *maximal output* $O_i$
    - Each plant manufactures at a *capacity-specific cost* $C_i(P_i)$, which gives the cost as a function of the production
    - Each customer $j = 1, \ldots, N$ *requests* $C_j$ goods
- **Objective function**: find the *optimal production schedule* such that the manufacturing and shipment costs are minimized

```{r echo=FALSE, fig.align="center", out.width="80%"}
include_figure("problem_product_scheduling.png")
```

### Portfolio Optimization
- **Problem**
    - Investor wants to *invest* money such that it *maximizes* the investor's *utility*
    - Utility $U$ depends on *daily return* $\mu$ and *risk* $\sigma^2$
    - Given risk taking $\kappa$, then $U(\mu, \sigma^2) = \mu - \frac{\kappa}{2} \sigma^2$
- **Objective function**: *Maximize $U(\mu, \sigma^2)$* among a range of stocks $s_1, \ldots s_N$

```{r echo=FALSE, fig.align="center", out.width="80%"}
df1 <- data.frame(
    "$\\mu$", 0.2, 0.5, 0.1
  )
colnames(df1) <- c("$\\vecval{c}^T$","$s_1$","$s_2$", "$s_3$")
df2 <- data.frame(
    c("$s_1$","$s_2$", "$s_3$"),
    c(0.1,0.02,0.02),
    c(0.02,0.1,0.02),
    c(0.02,0.02,0.1)
  )
colnames(df2) <- c("$Q$","$s_1$","$s_2$", "$s_3$")
knitr::kable(
  list(df1,df2),
  escape = FALSE,
  booktabs = TRUE
)
```


$$
\Rightarrow
\max\limits_{\vecval{x}} \vecval{c}^T \vecval{x} - \frac{\kappa}{2} \vecval{x}^T Q \vecval{x}
\quad
\text{s.t.}
\quad
x_i \geq 0 \text{ and } \sum\limits_i{x_i} = 1
$$

### Portfolio Optimization in R
```{r quadprog,eval=FALSE}
library(quadprog) # load necessary library
kappa <- 4 # set risk taking
# objective function
c <- c(0.02, 0.05, 0.01)
Q <- matrix(c(0.1,0.02,0.02, 0.02,0.1,0.02, 
              0.02,0.02,0.1), nrow=3)
# constraints
A <- matrix(c(1,1,0,0, 1,0,1,0, 1,0,0,1), nrow=4)
b <- c(1, 0, 0, 0)

sol <- solve.QP(kappa/2*Q, c, t(A), b, meq=1) # call solver
sol$solution # ratio of stocks in portfolio
sol$value # minimum value of objective function
```

## Outlook

1. *Introduction to R*
1. *Advanced R* 
    - Programming prerequisites 
    - Visualize optimization routines
1. *Numerical Analysis:* 
    - Mathematical prerequisites to derive and formalize optimization routines
1. *Optimization in R:*
    - Use of built-in optimization routines
