# Linear and Quadratic Optimization {#lp_qp}

## Linear Programming

Mathematical description

- In compact form

\begin{equation}
\min_{\vecval{x}} c^T \vecval{x} ~ = \min_{\vecval{x}} c_1x_1 + \dots + c_n x_n ~  \text{subject~to} ~ A\vecval{x} \geq \vecval{b} , ~ \vecval{x} \geq 0
\label{eq:lp:compact}
\end{equation}

- In matrix notation
\begin{equation}\begin{split}
\underbrace{
	\min_{\vecval{x}} 
		\begin{bmatrix}
			c_1 \\
			c_2 \\
			\vdots \\
			c_n
		\end{bmatrix}^T
		\begin{bmatrix}
			x_1 \\
			x_2 \\
			\vdots \\
			x_n
		\end{bmatrix}
		}_{\text{Objective}}
	\text{s.t.} ~
		\underbrace{
		\begin{bmatrix}
			a_{11} & a_{12} & \dots & a_{1n}\\
			a_{21} & a_{22} & \dots & a_{2n}\\
			\vdots & \vdots & \ddots & \vdots \\
			a_{m1} & a_{m2} & \dots & a_{mn}
		\end{bmatrix}
		\begin{bmatrix}
			x_1 \\
			x_2 \\
			\vdots \\
			x_n
		\end{bmatrix}
	\geq
		\begin{bmatrix}
			b_1 \\
			b_2 \\
			\vdots \\
			b_n
		\end{bmatrix}
		}_{\text{First constraint}}
	,~
	\underbrace{
		\begin{bmatrix}
			x_1 \\
			x_2 \\
			\vdots \\
			x_n
		\end{bmatrix}
	\geq 0
	}_{\text{Second constraint}}
\end{split}\label{eq:lp:matrix}\end{equation}

## Example Linear Programming

- *Problem description* 
    - Products A and B are sold at EUR~25 and EUR~20 respectively
	  - Product A requires 20 resources whereas product B needs 12
	  - Per day only 1800 resources are available
	  - Both products require 1/15~hours to be made
	  - We have a total of 8~hours time
- If $x_1$ is the number items produced of type A and $x_2$ the type B products, then the return or *objective function* is

\begin{equation}
\max_{\vecval{x}} 25 x_1 + 20 x_2 = \max_{\vecval{x}} \left[ \begin{array}{c} 25 \\ 20 \end{array} \right]^T \left[ \begin{array}{c} x_1 \\ x_2 \end{array} \right] 
\end{equation}

- The resource constraint is given by $20 x_1 + 12 x_2 \leq 1800$ and the hour constraint by $\frac{1}{15} x_1 + \frac{1}{15} x_2 \leq 8$

- Both constraints can be written in matrix form

\begin{equation}
A\vecval{x} =
\left[ \begin{array}{rr} 20 & 12 \\ \frac{1}{15}& \frac{1}{15} \end{array} \right] 
\left[ \begin{array}{r} x_1 \\ x_2 \end{array} \right]
\leq 
\left[ \begin{array}{r} 1800 \\ 8 \end{array} \right] = \vecval{b}
\end{equation}

- Visualization of the production problem and its constraints
```{r echo=FALSE,results='hide',fig.width=6,fig.height=4,out.width='80%',message=FALSE,warning=FALSE}
xy <- as.matrix(expand.grid(seq(0,75,length = 101),seq(25,150,length = 101)))
colnames(xy) <- c("x1", "x2")
df <- data.frame(xy)
df[["y"]] <- df$x1 * 25 + df$x2 * 20

#plot linear with wireframe
#require(lattice)
#wireframe(y ~ x1*x2, data = df, shade = TRUE, drape=FALSE, scales = list(arrows = FALSE), screen = list(z=-240, x=-70, y=0))

rx1 <- seq(0,75, length=101)
constraints <- data.frame(x = c(rx1, rx1), 
                          y=c((1800-20*rx1)/12, (8-1/15*rx1)*15), 
                          Constraint = c(rep("Resource", 101), rep("Time", 101)))

# plot linear as tiles
library(ggplot2)
print(ggplot(data=df, aes(x=x1, y=x2)) +
  geom_tile(aes(fill = y)) +
  geom_line(data = constraints, aes(x = x, y = y, group = Constraint, colour = Constraint), size = 2)+
  geom_point(aes(x=45, y=75), size = 4, colour = "black")+
  xlab("x_1") + ylab("x_2") +
  scale_fill_gradient(low="black", high="white", name="Objectve") +
  guides(fill=guide_colorbar(barwidth=15)) +
  theme_bw() +
  theme(legend.position="top"))
```

### Constrained Optimization

Describes the optimization of an *objective function* for a set of variables with equality or inequality constraints in place for these variables

## Linear Programming with R

- Coefficients $c$ are included in the vector `objective`.`in`, constraints $A$ given as matrix `const`.`mat`, constraint direction as `const`.`dir` and the constraint value $b$ as `const`.`rhs`
- In R package `optimize`

```{r eval=FALSE,size='scriptsize'}
require(lpSolve)
lp(direction="min", objective.in, const.mat, const.dir, 
const.rhs, int.vec)
```

- Implementation in R

```{r ,message=FALSE,warning=FALSE,size='scriptsize'}
require(lpSolve)
objective.in <- c(25, 20)
const.mat <- matrix(c(20, 12, 1/15, 1/15), nrow=2, byrow=TRUE)
const.rhs <- c(1800, 8)
const.dir <- c("<=", "<=")
optimum <- lp (direction = "max", objective.in, const.mat, 
const.dir, const.rhs)
optimum$solution
optimum$objval
```

## Quadratic Programming

Mathematical description

- In compact form

\begin{equation}
\min_{\vecval{x}} \frac{1}{2} \vecval{x}^T D \vecval{x} - \vecval{d}^T \vecval{x}  ~  \text{subject~to} ~ A^T\vecval{x} \geq \vecval{b} 
\label{eq:qp:compact}
\end{equation}

- In matrix notation

\begin{equation}
	\min_x 
		\begin{bmatrix}
			x_1 \\
			x_2 \\
			\vdots \\
			x_n
		\end{bmatrix}^T
		\begin{bmatrix}
			d_{11} & d_{12} & \dots & d_{1n}\\
			d_{21} & d_{22} & \dots & d_{2n}\\
			\vdots & \vdots & \ddots & \vdots \\
			d_{m1} & d_{m2} & \dots & d_{mn}
		\end{bmatrix}
		\begin{bmatrix}
			x_1 \\
			x_2 \\
			\vdots \\
			x_n
		\end{bmatrix}-
		\begin{bmatrix}
			d_1 \\
			d_2 \\
			\vdots \\
			d_n
		\end{bmatrix}
		\begin{bmatrix}
			x_1 \\
			x_2 \\
			\vdots \\
			x_n
		\end{bmatrix}
	\text{s.t.} ~
		\begin{bmatrix}
			a_{11} & a_{12} & \dots & a_{1n}\\
			a_{21} & a_{22} & \dots & a_{2n}\\
			\vdots & \vdots & \ddots & \vdots \\
			a_{m1} & a_{m2} & \dots & a_{mn}
		\end{bmatrix}^T
		\begin{bmatrix}
			x_1 \\
			x_2 \\
			\vdots \\
			x_n
		\end{bmatrix}
	\geq
		\begin{bmatrix}
			b_1 \\
			b_2 \\
			\vdots \\
			b_n
		\end{bmatrix}
\label{eq:qp:matrix}\end{equation}

## Quadratic Programming

- *Parameter mapping* in R
	  - *Quadratic coefficients* $D$ are mapped to `Dmat` 
    -  *Linear coefficients* $d$ are mapped to `dvec`
    - *Constraints matrix* $A$ is mapped to `Amat`
    - *Constraint equalities* or inequalities $b$ are provided in `bvec`
    - Parameter `meq`$=n$ sets the first $n$ entries as *equality* constraints; all further constraints are inequality
- Function call with package `quadprog`

```{r eval=FALSE}
require(quadprog)
solve.QP(Qmat, dvec, Amat, bvec, meq)
```

- Many problems can formulated in quadratic form, e.g., portfolio optimization, circus tent problem, demand response, ...

## Example Circus Tent

```{r setup,echo=FALSE,eval=TRUE}
library(knitr)
library(rgl)
knit_hooks$set(rgl = hook_rgl)
```

```{r circus_tent_problem,rgl=TRUE,fig.show='last',echo=FALSE,eval=TRUE,crop=TRUE,message=FALSE,warning=FALSE,results='hide',fig.width=4,fig.height=4,out.width='80%'}
# Circus tent
# Build the pole grid using basic symmetries.
# http://blog.ryanwalker.us/2014/04/the-circus-tent-problem-with-rs-quadprog.html
q2 = matrix(0,18,18)
q2[8:9,8:9] = .3
q2[17:18,17:18] = 1/2
q1 <- q2[,18:1]
top  <- cbind(q2,q1)
z <- rbind(top,top[18:1,])

# Plot
library(rgl)  
x <- (1:36)/36
y <- (1:36)/36
open3d()
rgl.surface(x,y, z, color='blue',alpha=.75, smooth=TRUE)
rgl.surface(x,y, matrix(1/2,36,36), color='red', front='lines', back='lines')
rgl.bg(color="white")
rgl.viewpoint(theta = 15, phi = 20, zoom = .8)
```

### Question

How to bring this into quadratic form?

- How to calculate the height of the tent at every point?
- Tent height at each grid point $(x, y)$ is given by $u(x, y)$
- Tent sheet settles into minimal energy state $E[u]$ for each height $u$
- Use the *Dirichlet energy* to estimate $E[u]$ of $u$
- We discretize the energy and ultimately come up with 

\begin{equation}
	E[u] \approx \frac{h_x h_y}{2} ~ \vecval{u}^T L \vecval{u}
\end{equation} 
		
- Which is quadratic

Full description:

http://blog.ryanwalker.us/2014/04/the-circus-tent-problem-with-rs-quadprog.html


### Visualized solution

```{r circus_tent_solution,rgl=TRUE,fig.show='last',echo=FALSE,eval=TRUE,crop=TRUE,message=FALSE,warning=FALSE,results='hide',fig.width=2,fig.height=2,out.width='80%'}
nr <- nrow(z)
nc <- ncol(z)
N  <- nr*nc
bvec <- matrix(z,N,1,byrow=FALSE)

# 2D Discrete Laplacian on an lx by ly rectangular grid with
# nx grid lines in the x direction and ny grid lines in the y direction.
Laplacian2D <- function(lx,ly, nx,ny){
  hx <- lx/(nx-1)
  hy <- ly/(ny-1)
  tr_x <- c(2,-1,rep(0,nx-2))
  tr_y <- c(2,-1,rep(0,ny-2))
  Tx <- toeplitz(tr_x)/(hx^2)
  Ty <- toeplitz(tr_y)/(hy^2)
  Ix <- diag(nx)
  Iy <- diag(ny)
  L <- kronecker(Tx,Iy) + kronecker(Ix,Ty)
  return(L)
}

# Solve the QP:
library(quadprog)
Ny  <- nrow(z)
Nx  <- ncol(z)
N   <- Nx*Ny
hx  <- 1/(Nx-1)
hy  <- 1/(Ny-1)

# System matrices
# Note: quadprog's quadratic form is -d^T x + 1/2 x^T D x with t(A)%*%x>= b
theta <- 1                            # coefficient for linear term
Dmat <- hx*hy*Laplacian2D(1,1,36,36)  
dvec <- -theta*(hx*hy)*rep(1,N)      
Amat <- t(diag(N))
bvec <- matrix(z,N,1,byrow=FALSE)     # lower bound

# call the solver
sol  <- solve.QP(Dmat, dvec, Amat, bvec)

# extract and reshape the solution
tent <- matrix(sol$solution,nrow(z),ncol(z), byrow=FALSE)

#plot  
open3d()
x <- (1:Nx)/Nx
y <- (1:Ny)/Ny
rgl.surface(x, y , z, color='blue',alpha=.75, smooth=TRUE)
rgl.surface(x, y , tent, color='red', front='lines', back='lines')
rgl.bg(color="white")
rgl.viewpoint(theta = 15, phi = 10, zoom = .7)
```

