# (APPENDIX) Control structures in R {#control_structures}

## Control Flow

### Managing Code Execution

- *Control flow* specifies order in which statements are executed
- Previous concepts can only execute R code in a *linear* fashion
- *Control flow constructs* can choose which execution path to follow

- Functions: Combines sequence of statements into a self-contained task
- Conditional expressions: Different computations according to a specific condition
- Loops: Sequence of statements which may be executed more than once




### Functions

- Functions *avoid repeating the same code* more than once 
- Leave the current evaluation context to execute pre-defined commands

```{r fig.align="center", fig.cap="", out.width="80%", echo=FALSE}
include_figure("function.png")
```

- Extend set of built-in functions with opportunity for customization
- Functions **can** consist of the following:
    1. *Name* to refer to (avoid existing function names in R)
    1. *Function body* is a sequence of statements
    1. *Arguments* define additional parameters passed to the function body
    1. *Return value* which can be used after executing the function
- Simple example:

```{r }
f <- function(x,y) {
  return(2*x + y^2)
}
f(-3, 5)
```

- General syntax

```{r eval=FALSE}
functionname <- function(argument1, argument2, ...) { 
  function_body
  return(value)
}
```

- *Return value* is the last evaluated expression
    - Alternative: set explicitly with `return(...)`
- Curly brackets can be omitted if the function contains only one statement (not recommended)
- Be cautious since the *order of the arguments matters*
- Values in functions are *not printed* in console\\
    - Remedy is `print(...)`



### Examples of Functions

```{r }
square <- function(x) x*x # last value is return value
square(10)
```

```{r }
cubic <- function(x) {
  # Print value to screen from inside the function
  print(c("Value: ", x, " Cubic: ", x*x*x))
  # no return value
}
cubic(10)
```

```{r }
hello <- function() { # no arguments
  print("world")
}
hello()
```

```{r } 
my.mean <- function(x) {
  return (sum(x)/length(x))
}
my.mean(1:100)
```

### Scope in Functions

- Variables created inside a function only exists within it $\rightarrow$ *local*
- They are thus inaccessible from outside of the function
- *Scope* denotes when the name binding of variable is valid

```{r }
x <- "A"
g <- function(x) {
  x <- "B"
  return(x)
}
x <- "C"
```

- What are the values?

```{r function_scope,eval=FALSE}
g(x) # Return value of function x
x    # Value of x after function execution
```

- Solution

```{r ,echo=FALSE}
<<function_scope>>
```

### Unevaluated Expressions

- Expressions can store *symbolic mathematical statements* for later modifications (e.g. symbolic derivatives)
- Let's define an example via `expression(...)`

```{r }
f <- expression(x^3+3*y-y^3-3*x)
f
```

- If *evaluation* of certain parameters becomes necessary, one can use `eval(...)`

```{r }
x <- 2
y <- 3
eval(f)
```

### If-Else Conditions

- *Conditional execution* requires a condition to be met

```{r fig.align="center", fig.cap="", out.width="80%", echo=FALSE}
include_figure("if-else.png")
```

### If-Else Conditions

- Keyword `if` with optional `else` clause
- General syntax:

**if condition**:

```{r ,eval=FALSE}
if (condition) {
  statement1
}
```

If `condition` is true, then `statement` is executed

**if-else condition**:

```{r ,eval=FALSE}
if (condition) {
  statement1
} else {
  statement2
}
```

If `condition` is true, then `statement1` is executed, otherwise `statement2`

### If-Else Conditions

- Example

```{r }
grade <- 2
if (grade <= 4) {
  print("Passed")
} else {
  print("Failed")
}
```

```{r }
grade <- 5
if (grade <= 4) {
  print("Passed")
} else {
  print("Failed")
}
```

- Condition must be of length 1 and evaluate as either `TRUE` or `FALSE`

```{r }
if (c(TRUE, FALSE)) { # don't do this!
  print("something")
}
```

### Else-If Clauses

- *Multiple conditions* can be checked with `else if` clauses
- The last `else` clause applies when no other conditions are fulfilled
- The same behavior can also be achieved with *nested if-clauses*

**else-if clause**:

```{r ,eval=FALSE,size='scriptsize'}
if (grade == 1) {
  print("very good")
} else if (grade == 2) {
  print("good")
} else {
  print("not a good grade")
}
```

**Nested if-condition**:

```{r ,eval=FALSE,size='scriptsize'}
if (grade == 1) {
  print("very good")
} else {
  if (grade == 2) {
    print("good")
  } else {
    print("not a good grade")
  }
}
```

### If-Else Function

- As an alternative, one can also reach the same control flow via the function `ifelse(...)` 

```{r ,eval=FALSE}
ifelse(condition, statement1, statement2) 
# executes statement1 if condition is true, 
# otherwise statement2
```

```{r }
grade <- 2
ifelse(grade <= 4, "Passed", "Failed")
```

- `ifelse(...)` can also work with vectors as if it was applied to each element separately

```{r }
grades <- c(1, 2, 3, 4, 5)
ifelse(grades <= 4, "Passed", "Failed")
```

- This allows for the efficient comparison of vectors

### For Loop

- `for` loops execute statements for a *fixed number of repetitions*

```{r fig.align="center", fig.cap="", out.width="80%", echo=FALSE}
include_figure("loop.png")
```

- General syntax

```{r eval=FALSE}
for (counter in looping_vector){
  # code to be executed for each element in the sequence
}
```

- In every iteration of the loop, one value in the looping vector is assigned to the `counter` variable that can be used in the statements of the body of the loop. 
- Examples

```{r eval=TRUE}
for (i in 4:7) {
  print(i)
}
```

```{r eval=TRUE}
a <- c()
for (i in 1:3){
  a[i] <- sqrt(i)
}
a
```

### While Loop

- *Loop* where the number of iterations is *controlled by a condition*
- The condition is checked in every iteration
- When the condition is met, the loop body in curly brackets is executed 
- General syntax

```{r eval=FALSE}
while (condition) {
  # code to be executed
}
```

- Examples

```{r ,size='scriptsize'} 
z <- 1
# same behavior as for loop
while (z <= 4) { 
    print(z)  
    z <- z + 1
}
```

```{r ,size='scriptsize'} 
z <- 1
# iterates all odd numbers
while (z <= 5) { 
    z <- z + 2
    print(z)  
}
```

## Timing

### Measuring Timings via Stopwatch

- *Efficiency* is a major issue with larger datasets and complex codes
- Timings can help in understanding scalability and bottlenecks 
- Use a stopwatch approach measuring the duration between two `proc.time()` calls

```{r ,size='scriptsize'}  
start.time <- proc.time() # Start the clock

g <- rnorm(100000)
h <- rep(NA, 100000)
for (i in 1:100000) { # Loop over vector, always add +1
  h[i] <- g[i] + 1
}

# Stop clock and measure duration
duration <- proc.time() - start.time
```

### Measuring Timings via Stopwatch

- Results of `duration` have the following format

```{r ,echo=FALSE}
duration
```

- Timings are generally grouped into 3 categories
    - *User* time measures the understanding of the R instructions
    - *System* time measures the underlying execution time
    - *Elapsed* is the difference since starting the stopwatch ($=$ user $+$ system)
- Alternative approach avoiding loop

```{r }
start.time <- proc.time() # Start clock
g <- rnorm(100000)
h <- g + 1
proc.time() - start.time # Stop clock
```

- Rule: *vector operations are faster* than loops

### Measuring Timings of Function Calls

Function `system.time(...)` can directly time function calls

```{r }
slowloop <- function(v){ 
	for (i in v) {
	  tmp <- sqrt(i)
	}
}

system.time(slowloop(1:1000000)) 
```
