# (APPENDIX) Visualization in R {#visualization}

## Point Plot

- Creating simple point plots (also named scatter plots) via `plot(...)`
- Relies upon vectors denoting the x-axis and y-axis locations
- Various options can be added to change appearance

```{r pointplot,out.width='\\textwidth',dpi=50,echo=-1,tidy=FALSE,size='scriptsize'}
setwd("../../Exercise/Datasets/AdvancedR/")
d <- read.csv("persons.csv", header=TRUE, sep=",", 
              stringsAsFactors=FALSE)
plot(d$height, d$age)
```

## Adding Title, Labels and Annotations

- Title is added through additional parameter `main`
- Axis labels are set via `xlab` and `ylab`
- Annotations next to points with `text(...)`

```{r pointplot-with-labels,out.width='.8\\textwidth',fig.keep='last',dpi=50,tidy=FALSE}
plot(d$height, d$age, 
     main="Title",              # title for the plot 
     xlab="Height", ylab="Age") # labels for x and y axis
text(d$height, d$age, d$name) # d$name are annotations
```

## Line Plot

Generate line plot using the additional option `type="l"`

```{r lineplot,out.width='\\textwidth',fig.keep='last',dpi=50}
x <- seq(0, 4, 0.01)
plot(x, x*x, type="l")
```

## Exercise: Plotting

```{r exercise_plotting_x,eval=FALSE}
x <- seq(-1, +1, 0.01)
```

```{r ,include=FALSE}
<<exercise_plotting_x>>
```

```{r exercise_plotting,out.width='.5\\textwidth',fig.keep='last',dpi=50,echo=FALSE}
plot(x, abs(x), type="l", ylab="", xlab="")
```

### Question
- How would you reproduce the above plot?
    - `plot(x, kink(x), type="l", main="")`
    - `plot(x, kink(x), type="l", lab="")`
    - `plot(x, abs(x), type="l", ylab="", xlab="")`
- Pingo? TODO

## 3D Plots

- Consider the function $f(x,y) = x^3 + 3y - y^3 - 3x$

```{r }
f <- function(x, y) x^3+3*y-y^3-3*x
```

- Create axis ranges for plotting

```{r }
x <- seq(-5, 5, 0.1)
y <- seq(-5, 5, 0.1)
```

- Function `outer(x,y,f)` evaluates `f` all combinations of `x` and `y`

```{r }
z <- outer(x, y, f)
```

## 3D Plots

Function `persp(...)` plots the plane through `x`, `y` and `z` in 3D

```{r persp_plot,size='scriptsize',out.width='.7\\textwidth',fig.width=6,fig.height=6}
persp(x, y, z)
```

Turn on ticks on axes via `ticktype="detailed"`
```{r persp_plot_parameters,size='scriptsize',out.width='.7\\textwidth',fig.width=6,fig.height=6}
persp(x, y, z, ticktype="detailed")
```

Parameters `theta` (left/right) and `phi` (up/down) control viewing angle

```{r persp_plot_parameters1,size='scriptsize',out.width='\\textwidth',fig.width=4,fig.height=4}
persp(x, y, z, theta=20, phi=0)
```

```{r persp_plot_parameters2,size='scriptsize',out.width='\\textwidth',fig.width=4,fig.height=4}
persp(x, y, z, theta=20, phi=35)
```

## Contour Plots

- A *contour line* is a curve along which the function has the same value
- `image(...)` plots a grid of pixels colored corresponding to `z`-value
- `contour(..., add=TRUE)` adds contour lines to an existing plot

```{r contour_plot,size='scriptsize',fig.width=6,fig.height=4}
image(x, y, z) # Plot colors
contour(x, y, z, add=TRUE) # Add contour lines
```

```{r contour_plot_sphere,size='scriptsize',eval=FALSE}
f <- function(x, y) sqrt(x^2+y^2)
z <- outer(x, y, f)
image(x, y, z, asp=1) # set aspect ratio, i.e. same scale for x and y
contour(x, y, z, add=TRUE)
```

### Question

- What would the above plot look like?

```{r ,echo=FALSE,out.width='50%',fig.width=3,fig.height=3, fig.cap="Answer A"}
f <- function(x, y) sqrt(10*x^2+y^2)
z <- outer(x, y, f)
image(x, y, z, asp=1)
contour(x, y, z, add=TRUE)
```
 
```{r ,echo=FALSE,out.width='50%',fig.width=3,fig.height=3, fig.cap="Answer B"}
<<contour_plot_sphere>>
```

```{r ,echo=FALSE,out.width='50%',fig.width=3,fig.height=3, fig.cap="Answer C"}
f <- function(x, y) sin(x/4)
z <- outer(x, y, f)
image(x, y, z, asp=1)
contour(x, y, z, add=TRUE)
```

- Pingo? TODO

## Plotting Regression Plane

```{r size='tiny',tidy=FALSE}
library(car) # for dataset Highway1
model <- lm(rate ~ len + slim, data=Highway1)
model

x1r <- range(Highway1$len)
x1seq <- seq(x1r[1], x1r[2], length=30)
x2r <- range(Highway1$slim)
x2seq <- seq(x2r[1],x2r[2], length=30)
z <- outer(x1seq, x2seq, 
           function(a,b) predict(model, 
                                 newdata=data.frame(len=a,slim=b)))
```

```{r ,fig.width=6,fig.height=4,tidy=FALSE}
res <- persp(x=x1seq, y=x2seq, z=z,
             theta=50, phi=-10)
dp <- trans3d(Highway1$len, Highway1$slim, 
                    Highway1$rate, pmat=res)
points(dp, pch=20, col="red")
```
